---
title: "Fine-mapping a 2df joint test with SuSiEgxe"
output:
  rmarkdown::html_vignette:
    dev: !expr if (isTRUE(capabilities("cairo"))) "svg" else "pdf"
vignette: >
  %\VignetteIndexEntry{Fine-mapping a 2df joint test with SuSiEgxe}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6, fig.height = 3.5)
```

## Overview

SuSiEgxe extends the SuSiE ("sum of single effects") fine-mapping model to
summary statistics from the **2df joint test** in gene-by-environment (GxE)
interaction meta-analysis with a **binary exposure**. For each variant it uses
the main-effect estimate, the interaction-effect estimate, their standard
errors, and the covariance between the two estimates, together with an LD
matrix. Up to `L` independent signals can be fine-mapped in one locus, and each
signal gets its own credible set.

This vignette walks through one complete analysis using the example data that
ships with the package: diastolic blood pressure (DBP) and current smoking in an
East Asian ancestry (EAS) meta-analysis, locus `2:26681997-26710019`.

This vignette was built with SuSiEgxe version `r packageVersion("SuSiEgxe")`.
See the change log at the end for what changed between versions. Results
produced with versions before 0.2.0 should be regenerated; see the entry for
version 0.2.0.

```{r setup}
library(SuSiEgxe)
```

## Requirements on the input

Three things must hold for the results to be meaningful.

1. **The exposure is coded 0/1 and is not centered.** The way the package
   rebuilds the sufficient statistics from summary statistics is derived for
   this coding. With it, the covariance between the main-effect and the
   interaction-effect estimates is negative at every variant, and
   `susie_rss_gxe()` stops if any value of `covhat` is zero or positive. A
   negative covariance is necessary but does not by itself prove that the
   exposure is 0/1.
2. **The LD matrix is a signed correlation matrix** (r, not r-squared and not
   the absolute value of r), computed with the same effect-allele coding as the
   summary statistics. When more than one signal is fitted, the sign of the
   correlation between variants is used, and an unsigned matrix gives wrong
   results.
3. **The summary statistics and the LD matrix describe the same variants in
   the same order**, on the same effect allele.

In a meta-analysis, variants in one locus can have very different sample sizes
and the LD matrix usually comes from a reference panel. Both can make a
multi-signal fit unstable. The section "Check convergence" shows what to look
at.

## Input data

Two files are needed: a table of summary statistics and an LD matrix whose row
and column names are variant IDs.

```{r data}
sumstat_file <- system.file("extdata",
  "EAS.DBP.CURSMK.COMBINED.2df.2_26681997_26710019.rsid.txt",
  package = "SuSiEgxe")
ld_file <- system.file("extdata",
  "EAS.DBP.CURSMK.COMBINED.2df.2_26681997_26710019.LD_mat.txt",
  package = "SuSiEgxe")

summstats <- read.delim(sumstat_file, stringsAsFactors = FALSE)
ld_matrix <- as.matrix(read.table(ld_file, header = TRUE))

dim(summstats)
dim(ld_matrix)
```

The columns used by the model are:

| Column       | Meaning                                              |
|--------------|------------------------------------------------------|
| `Effect`     | Main (genetic) effect estimate                       |
| `StdErr`     | Standard error of the main effect                    |
| `IntEffect`  | Interaction (GxE) effect estimate                    |
| `IntStdErr`  | Standard error of the interaction effect             |
| `IntCov`     | Covariance of the main and interaction estimates     |
| `refsnp_id`  | Variant ID, matching the LD matrix names             |

## Align statistics and LD

The summary statistics and the LD matrix must list the same variants in the
same order.

```{r align}
summstats <- summstats[summstats$refsnp_id %in% colnames(ld_matrix), ]
idx <- match(summstats$refsnp_id, colnames(ld_matrix))
R <- ld_matrix[idx, idx]

stopifnot(nrow(summstats) == nrow(R))
dim(R)
```

## Check the input

Check the two requirements that can be read off the data.

```{r check-input}
# 0/1 exposure: the covariance must be negative at every variant.
all(summstats$IntCov < 0)

# Under 0/1 coding the covariance is close to minus the main-effect variance,
# so this ratio should be close to 1.
summary(-summstats$IntCov / summstats$StdErr^2)

# Signed LD: a signed correlation matrix has negative entries. A minimum of 0
# means the matrix is r-squared or |r| and must be replaced before fitting
# more than one signal.
range(R[upper.tri(R)])
```

## Run the model

`susie_rss_gxe()` takes the five summary-statistic vectors and `R`. The prior
for each single effect is a 2x2 matrix: the prior variance of the main effect
and of the interaction effect on the diagonal, and their prior covariance off
the diagonal. Here we use the same prior for each of `L = 5` effects and keep it
fixed. `prior_variance` can be given as a single 2x2 matrix, which is then used
for every effect, or as a list of `L` matrices.

```{r fit}
L <- 5
prior_var <- matrix(c(10, -0.02, -0.02, 10), ncol = 2)

fit <- susie_rss_gxe(
  R = R,
  bhat = summstats$Effect,
  bhat_gxe = summstats$IntEffect,
  shat = summstats$StdErr,
  shat_gxe = summstats$IntStdErr,
  covhat = summstats$IntCov,
  z_ld_weight = 0,
  estimate_residual_variance = FALSE,
  prior_variance = prior_var,
  estimate_prior_variance = FALSE,
  check_prior = TRUE,
  L = L,
  max_iter = 600,
  min_abs_corr = 0,
  verbose = FALSE
)
```

Notes on the arguments:

- `L` is the maximum number of signals. Effects that find no signal do not
  produce a credible set. Setting `L = 1` gives single-signal fine-mapping,
  which does not use the off-diagonal entries of `R`.
- `estimate_residual_variance` must be `FALSE`. The residual variance is fixed
  and absorbed into the summary statistics, and the function stops if it is set
  to `TRUE`.
- `estimate_prior_variance = TRUE` estimates the 2x2 prior of each effect from
  the data instead of keeping it fixed.
- No sample size `n` is supplied, so the model assumes a large sample and small
  effects (the package prints a warning to that effect).

## Check convergence

By default the algorithm stops when the objective (ELBO) increases by less than
`tol` between two iterations, as in susieR. Use `convergence_method = "pip"` to
stop when no posterior inclusion probability changes by more than `tol`.

```{r convergence}
fit$converged
fit$niter

# The objective should never decrease.
all(diff(fit$elbo) > -1e-6)
```

Treat the fit with caution if any of the following happens:

- `fit$converged` is `FALSE` after `max_iter` iterations.
- The package warns that the objective decreased.
- The objective keeps growing by large amounts from one iteration to the next.
- Several credible sets contain a single variant each, and those variants are
  in high LD with one another.

These usually mean that the summary statistics and the LD matrix are not
consistent with each other: an unsigned or misaligned LD matrix, a reference
panel that does not match the study population, or variants with very
different sample sizes within the locus.

## Posterior inclusion probabilities

```{r pip}
pip <- susie_get_pip(fit)
names(pip) <- summstats$refsnp_id
head(sort(pip, decreasing = TRUE), 8)
```

```{r pip-plot, eval = isTRUE(capabilities("cairo"))}
plot(pip, pch = 16, xlab = "Variant index", ylab = "PIP",
     main = "Posterior inclusion probabilities", ylim = c(0, 1))
```

## Credible sets

`susie_get_cs()` extracts credible sets. Passing the LD matrix through `Xcorr`
lets it compute the purity of each set.

```{r cs}
cs <- susie_get_cs(fit, Xcorr = R, coverage = 0.95, min_abs_corr = 0)

cs$cs
cs$purity
cs$coverage
```

Within one credible set the posterior inclusion probabilities should add up to
about the coverage, not to more than 1:

```{r cs-pip-sum}
sapply(cs$cs, function(i) sum(pip[i]))
```

Map the variant indices in a credible set back to variant IDs:

```{r cs-ids}
summstats$refsnp_id[cs$cs[[1]]]
```

## Running your own locus

1. Subset your summary statistics and LD matrix to one locus.
2. Make sure the exposure is coded 0/1 and not centered, and that the LD matrix
   is a signed correlation matrix on the same effect-allele coding as the
   summary statistics (see "Requirements on the input").
3. Put both in the same variant order (see "Align statistics and LD") and run
   the checks in "Check the input".
4. Choose `L`, the maximum number of causal signals, and a 2x2 `prior_variance`.
5. Call `susie_rss_gxe()`, check convergence, then use `susie_get_pip()` and
   `susie_get_cs()`.

The package also exports `susie_suff_stat_gxe()`, which accepts sufficient
statistics (`XtX`, `XtZ`, `ZtZ`, `Xty`, ...) directly; see `?susie_suff_stat_gxe`.

## Session information

```{r session-info}
sessionInfo()
```

## Version

The current version is 0.2.0 (Oct 3, 2026).

### Change log

**Version 0.2.0 (Oct 3, 2026)**

This version fixes errors that affect results. Analyses run with earlier
versions should be rerun.

Bug fixes that change results:

- Each single-effect update now uses the residual, after the other effects are
  removed, to compute the Bayes factors, the inclusion probabilities and the
  prior update. The credible set returned by earlier versions is the
  single-signal (`L = 1`) credible set. We thank Yihe Yang, Feng Yan and Josée
  Dupuis for reporting this.

Input checks and unsupported options:

- `susie_rss_gxe()` stops if any `covhat` is zero or positive. The package
  supports an uncentered 0/1 exposure only.
- `estimate_residual_variance = TRUE`, `estimate_prior_method = "EM"`, `s_init`
  and `refine = TRUE` now stop with a message instead of calling undefined
  functions.

Other changes:

- Convergence is declared on the objective (ELBO) by default, as in susieR.
  The previous criterion is available with `convergence_method = "pip"`. A
  warning is given if the objective decreases.
- `prior_variance` can be a single 2x2 matrix, so the default call works.
- The log Bayes factors are computed in closed form for all variants at once,
  which makes `estimate_prior_variance = TRUE` much faster.
- `track_fit = TRUE` now works.

**Version 0.1.1 (Oct 3, 2026)**

- Added this vignette.
- Fixed a missing import of `Matrix::Diagonal()`. The model no longer requires
  `library(Matrix)` to be attached first.

**Version 0.1.0 (Feb 18, 2025)**

- First release. `susie_rss_gxe()` and `susie_suff_stat_gxe()` fine-map
  summary statistics from the 2df joint test in GEI meta-analysis with a binary
  exposure.
