---
title: "Fine-mapping a 2df joint test with SuSiEgxe"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fine-mapping a 2df joint test with SuSiEgxe}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6, fig.height = 3.5)
```

## Overview

SuSiEgxe extends the SuSiE ("sum of single effects") fine-mapping model to
summary statistics from the **2df joint test** in gene-by-environment (GxE)
interaction meta-analysis with a **binary exposure**. For each variant it uses
the main-effect estimate, the interaction-effect estimate, their standard
errors, and the covariance between the two estimates, together with an LD
matrix.

This vignette walks through one complete analysis using the example data that
ships with the package: diastolic blood pressure (DBP) and current smoking in an
East Asian ancestry (EAS) meta-analysis, locus `2:26681997-26710019`.

```{r setup}
library(SuSiEgxe)
```

## Input data

Two files are needed: a table of summary statistics and an LD matrix whose row
and column names are variant IDs.

```{r data}
sumstat_file <- system.file("extdata",
  "EAS.DBP.CURSMK.COMBINED.2df.2_26681997_26710019.rsid.txt",
  package = "SuSiEgxe")
ld_file <- system.file("extdata",
  "EAS.DBP.CURSMK.COMBINED.2df.2_26681997_26710019.LD_mat.txt",
  package = "SuSiEgxe")

summstats <- read.delim(sumstat_file, stringsAsFactors = FALSE)
ld_matrix <- as.matrix(read.table(ld_file, header = TRUE))

dim(summstats)
dim(ld_matrix)
```

The columns used by the model are:

| Column       | Meaning                                              |
|--------------|------------------------------------------------------|
| `Effect`     | Main (genetic) effect estimate                       |
| `StdErr`     | Standard error of the main effect                    |
| `IntEffect`  | Interaction (GxE) effect estimate                    |
| `IntStdErr`  | Standard error of the interaction effect             |
| `IntCov`     | Covariance of the main and interaction estimates     |
| `refsnp_id`  | Variant ID, matching the LD matrix names             |

## Align statistics and LD

The summary statistics and the LD matrix must list the same variants in the
same order.

```{r align}
summstats <- summstats[summstats$refsnp_id %in% colnames(ld_matrix), ]
idx <- match(summstats$refsnp_id, colnames(ld_matrix))
R <- ld_matrix[idx, idx]

stopifnot(nrow(summstats) == nrow(R))
dim(R)
```

## Run the model

`susie_rss_gxe()` takes the five summary-statistic vectors and `R`. The prior
for each single effect is a 2x2 matrix: the prior variance of the main effect
and of the interaction effect on the diagonal, and their prior covariance off
the diagonal. Here we use the same prior for each of `L = 5` effects and keep it
fixed.

```{r fit}
L <- 5
prior_var <- replicate(L, matrix(c(10, -0.02, -0.02, 10), ncol = 2),
                       simplify = FALSE)

fit <- susie_rss_gxe(
  R = R,
  bhat = summstats$Effect,
  bhat_gxe = summstats$IntEffect,
  shat = summstats$StdErr,
  shat_gxe = summstats$IntStdErr,
  covhat = summstats$IntCov,
  z_ld_weight = 0,
  estimate_residual_variance = FALSE,
  prior_variance = prior_var,
  estimate_prior_variance = FALSE,
  check_prior = TRUE,
  L = L,
  max_iter = 600,
  min_abs_corr = 0,
  verbose = FALSE
)

fit$converged
fit$niter
```

No sample size `n` is supplied, so the model assumes a large sample and small
effects (the package prints a warning to that effect).

## Posterior inclusion probabilities

```{r pip}
pip <- susie_get_pip(fit)
names(pip) <- summstats$refsnp_id
head(sort(pip, decreasing = TRUE), 8)
```

```{r pip-plot}
plot(pip, pch = 16, xlab = "Variant index", ylab = "PIP",
     main = "Posterior inclusion probabilities", ylim = c(0, 1))
```

## Credible sets

`susie_get_cs()` extracts credible sets. Passing the LD matrix through `Xcorr`
lets it compute the purity of each set.

```{r cs}
cs <- susie_get_cs(fit, Xcorr = R, coverage = 0.95, min_abs_corr = 0)

cs$cs
cs$purity
cs$coverage
```

Map the variant indices in a credible set back to variant IDs:

```{r cs-ids}
summstats$refsnp_id[cs$cs[[1]]]
```

## Running your own locus

1. Subset your summary statistics and LD matrix to one locus.
2. Make sure both are on the same effect-allele coding and in the same variant
   order (see "Align statistics and LD").
3. Choose `L`, the maximum number of causal signals, and a 2x2 `prior_variance`.
4. Call `susie_rss_gxe()`, then `susie_get_pip()` and `susie_get_cs()`.

The package also exports `susie_suff_stat_gxe()`, which accepts sufficient
statistics (`XtX`, `XtZ`, `ZtZ`, `Xty`, ...) directly; see `?susie_suff_stat_gxe`.
